---
title: "final_project_code"
format: html
editor: visual
---

##Isha's Code - Excess Deaths and Pneumonia Trend Analysis (Hidden Burden of Covid-19)

---
title: "COVID-19 Excess Deaths Analysis"
format: html
editor: visual
---

## Data

```{r}
library(httr2)
library(dplyr)
library(lubridate)
library (janitor)
library(ggplot2)
library(scales)
library(knitr)
library(tidyverse)

#getwd()

##Retreving Data from cdc Website for covid cases, deaths, hospitalizations and ##vaccines

get_cdc_data_v <- function(endpoint) {
  api_url <- paste0("https://data.cdc.gov/resource/", endpoint, ".json")
  
  data <- request(api_url) |> 
    req_url_query("$limit" = 10000000) |> 
    req_perform() |> 
    resp_body_json(simplifyVector = TRUE)
  
  return(data)
}
#cases_raw <- get_cdc_data("pwn4-m3yp")
#hosp_raw <- get_cdc_data("39z2-9zu6")
deaths_raw_v <- get_cdc_data_v("r8kw-7aab")
#vax_raw <- get_cdc_data("rh2h-3yt2")
#head(deaths_raw_v)



# Working on the covid deaths data 2020-2025
#View(deaths_raw)
#head(deaths_raw_v)
state_abbreviations_v <- c(state.abb, "DC", "PR")
state_names_v <- c(state.name, "District of Columbia", "Puerto Rico")
# Convert and summarize deaths data
deaths_clean_v <- deaths_raw_v %>%
  filter(state != "United States")%>%
  mutate(
    start_date = as_date(start_date),
    covid_19_deaths = as.numeric(covid_19_deaths),
    mmwr_year = epiyear(start_date),
    mmwr_week = epiweek(start_date),
  )%>%
  select(start_date, state, mmwr_year, mmwr_week, covid_19_deaths) %>%
  filter(start_date > as.Date("2020-01-01")) %>%
  arrange(state, mmwr_year)  
#print(head(deaths_clean_v))



library(zoo)  

# Summarize deaths per day (national total)
us_deaths_daily_v <- deaths_clean_v %>%
  group_by(start_date) %>%
  summarise(total_deaths = sum(covid_19_deaths, na.rm = TRUE), .groups = "drop") %>%
  arrange(start_date) %>%
  mutate(rolling_avg_7day = zoo::rollmean(total_deaths, k = 7, fill = NA, align = "right"))

# Plot
plot1_v <- ggplot(us_deaths_daily_v, aes(x = start_date)) +
  geom_col(aes(y = total_deaths), fill = "red", alpha = 0.8) +
  geom_line(aes(y = rolling_avg_7day), color = "darkred", size = 1) +
  labs(
    title = "Daily COVID-19 Deaths in the US",
    subtitle = "With 10-Day Rolling Average",
    x = "Date",
    y = "Number of Deaths"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))+
  scale_x_date(date_breaks = "6 months",date_labels = "%b %Y")+
  scale_y_continuous(limits = c(0, 150000), labels = scales::label_comma())
#plot1_v

#ggsave("docs/new/Covid deaths in US (2020-2025).pdf", plot= plot1, width = 12, height = 8)
#ggsave("docs/new/Covid deaths in US (2020-2025).png", plot = plot1, width = 12, height = 8, dpi = 300)


```

```{r}
library(ggplot2)
library(scales)
library(dplyr)
library(lubridate)

## Divide the pandemic period, January 2020 to April 2025 into waves

variant_waves_v <- tibble::tibble(
  start = as_date(c("2020-03-01", "2021-01-01", "2021-07-01", "2021-12-01", "2022-04-01", "2023-06-15")),
  end   = as_date(c("2021-02-28", "2021-06-30", "2021-10-30", "2022-03-31", "2023-06-14", "2024-03-31")),
  variant = c("1st Wave: Original", 
              "2nd Wave: Alpha", 
              "3rd Wave: Delta", 
              "4th Wave: Omicron BA.1", 
              "5th Wave: Omicron Other",
              "6th Wave: XBB")
)

# Plot2: Covid deaths with variant overlay by state
plot2_v <- ggplot(us_deaths_daily_v, aes(x = start_date, y = total_deaths)) +
  geom_rect(
    data = variant_waves_v,
    aes(xmin = start, xmax = end, ymin = -Inf, ymax = Inf, fill = variant),
    inherit.aes = FALSE,
    alpha = 0.2
  ) +
  geom_line(aes(y = rolling_avg_7day), color = "darkred", size = 1) +
  scale_x_date(date_labels = "%b/%Y", date_breaks = "6 months") +
  labs(
    title = "US Weekly COVID-19 Deaths with Variant Waves (2020–2025)",
    x = "Month/Year",
    y = "Total Deaths",
    fill = "Variant Waves"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))+
  scale_y_continuous(limits = c(0, 150000), labels = scales::label_comma())
#plot2_v
#ggsave("docs/Covid deaths with variant overlay by state.pdf", plot= plot2, width = 12, height = 8)
#ggsave("docs/Covid deaths with variant overlay by state.png", plot = plot2, width = 12, height = 8, dpi = 300)
```

```{r}
library(dplyr)
library(lubridate)
library(stringr)
library(ggplot2)
library(viridis)

##A heatmap showing intensity of deaths by week and state.

deaths_filtered_v <- deaths_clean_v %>%
  filter(!state %in% c("United States", "New York City"))
deaths_quarterly_v <- deaths_filtered_v %>%
  mutate(quarter = paste0(year(start_date), " Q", quarter(start_date))) %>%
  group_by(state, quarter) %>%
  summarise(total_deaths = sum(covid_19_deaths, na.rm = TRUE), .groups = "drop")
deaths_quarterly_v$quarter <- factor(deaths_quarterly_v$quarter, levels = unique(deaths_quarterly_v$quarter))

#plot4: Heatmap showing intensity of deaths by state
plot4_v <- ggplot(deaths_quarterly_v, aes(x = quarter, y = state, fill = total_deaths)) +
  geom_tile(color = "white") +
  scale_fill_viridis(option = "magma", name = "Deaths", direction = -1) +
  labs(
    title = "COVID-19 Deaths by State (Quarterly)",
    x = "Quarter",
    y = "State"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 7),
    axis.text.y = element_text(size = 6)
  )
#plot4_v  
#ggsave("docs/new/COVID-19 Deaths by State (Quarterly).pdf", plot= plot4, width = 12, height = 8)
#2ggsave("docs/new/COVID-19 Deaths by State (Quarterly).png", plot = plot4, width = 12, height = 8, dpi = 300)
```

```{r}
## Reteriving the population of US states form Census website


library(jsonlite)
library(purrr)
library(tidyverse)
census_key_v <- "a498a5717131540bcea2c82b3e8fd47a367cef98"
url_v <- "https://api.census.gov/data/2021/pep/population"
request_v <- request(paste0(url_v,
    paste0("?get=POP_2020,POP_2021,NAME&for=state:*&key=",
           census_key_v)))
#print(request_v)
response_v <- request_v |> req_perform() 
#response_v

#content_type_v <- resp_content_type(response_v)
#cat("The content type of the response is:", content_type_v, "\n")
population_v <- response_v |>resp_body_string()|>jsonlite::fromJSON()
#head(population_v)

population_v <- population_v %>%
  row_to_names(row_number = 1)%>%
  as_tibble()
#head(population_v)

pop_latest_v<- read.csv("C:/Users/indhi/Documents/final_project_620/data/population_estimate_census.csv")
#head(pop_latest)
pop_latest_new2_v <- pop_latest_v %>%
  mutate(
    POP_2022 = as.character(gsub(",", "", X2022)),
    POP_2023 = as.character(gsub(",", "", X2023)),
    POP_2024 = as.character(gsub(",", "", X2024)),
    state_name = str_remove(state_name, "^\\.")
  ) %>%
  rename(NAME= "state_name") %>% 
  select(NAME, POP_2022, POP_2023, POP_2024) %>% 
  as.data.frame()
#pop_latest_new2_v 

population_new_v <- population_v %>% 
  left_join(pop_latest_new2_v, by = "NAME")
#head(population_new_v)

cleaned_population_v <- population_new_v %>% #changing this
  select(-state) %>%
  rename(state_name = NAME)%>%
  pivot_longer(cols = starts_with("POP_"),  
               names_to = "Year",
               values_to = "Population") %>%
  mutate(Year = gsub("POP_", "", Year),  
         Year = as.integer(Year), 
         Population = as.numeric(Population))  

state_abbreviations_v <- c(state.abb, "DC", "PR")
state_names_v <- c(state.name, "District of Columbia", "Puerto Rico")

cleaned_population_v <- cleaned_population_v %>%
  mutate(
    state = case_when(
      state_name %in% state_names_v ~ state_abbreviations_v[match(state_name, state_names_v)]
    )
  )
#head(cleaned_population_v)
#unique(cleaned_population_v$Year)
#colnames(cleaned_population_v)

#adding Region name data
url_iv <- "https://github.com/datasciencelabs/2024/raw/refs/heads/main/data/regions.json"
# regions <- use fromJSON to read as a data.frame
regions_data_v <- fromJSON(url_iv)
#str(regions_data)
#head(regions_data)
#View(regions_data)
#unique(regions_data_v$region_name)
# Convert the JSON data to a data frame and process it
regions_v <- regions_data_v %>%
  # Unnest the states from the nested list
  unnest(cols = states) %>%
  # Rename the columns
  rename(state_name = states) %>%
  # Rename the region with a long name to something shorter
  mutate(region_name = ifelse(region_name == "New York and New Jersey, Puerto Rico, Virgin Islands", "NYNJPV", region_name)) %>%
  # Ensure the region_name column is a factor
  mutate(region_name = factor(region_name))

# Print the first few rows of the regions data frame
#head(regions_v)
#view(regions)
#class(regions$region_name)

cleaned_population_v <- cleaned_population_v %>%
  left_join(regions_v, by = "state_name")
#head(cleaned_population_v)

all_dates_v <- data.frame(date = seq(make_date(2020, 1, 25),
                                   make_date(2024, 12, 31), 
                                   by = "week")) |>
  mutate(date = ceiling_date(date, unit = "week", week_start = 7) - days(1)) |>
  mutate(mmwr_year = epiyear(date), mmwr_week = epiweek(date)) 


dates_and_pop_v <- cross_join(all_dates_v, data.frame(state = unique(cleaned_population_v$state))) |> left_join(cleaned_population_v, by = c("state", "mmwr_year" = "Year"))
#head(dates_and_pop_v)

#unique(dates_and_pop_v$mmwr_year)
#View(dates_and_pop)


dates_and_pop1_v <-dates_and_pop_v %>% 
  mutate(mmwr_year= as.character(mmwr_year)) 
  
#head(dates_and_pop1_v)
#colnames(deaths_clean_v)
#unique(deaths_clean_v$mmwr_year)

us_deaths_weekly_v <- deaths_clean_v %>%
  rename(state_name= "state")%>%
  group_by(start_date,state_name, mmwr_week, mmwr_year) %>%
  summarise(total_deaths = sum(covid_19_deaths, na.rm = TRUE), .groups = "drop") %>% 
  arrange(state_name, mmwr_week)
#unique(us_deaths_weekly_v$mmwr_year)

us_deaths_weeklyregion_v <-us_deaths_weekly_v %>% 
  left_join(regions_v, by = "state_name")
#head(us_deaths_weeklyregion_v)  

#View(us_deaths_weeklyregion)
# merging the population data
datdeaths_v<-us_deaths_weeklyregion_v %>%
  left_join(dates_and_pop_v, by = c("state_name", "mmwr_year", "mmwr_week","region_name"))

#View(datdeaths1)

#datdeaths <- dates_and_pop %>%
  #left_join(us_deaths_weekly, by = c("state_name", "mmwr_year", "mmwr_week"))  
  #left_join(pop_latest_new1,by = c("state_name", "mmwr_year","mmwr_week"))
#head(datdeaths_v)
#View(datdeaths_v)

#colnames(datdeaths_v)
#unique(datdeaths_v$mmwr_year)

```

```{r}
##  For each period, compute the death rates by state. Describe which states did 
## better or worse during the different periods.


#  Merge variant period info to deaths+population dataset
datdeaths_waves_v <- datdeaths_v %>%
  mutate(wave = case_when(
    start_date >= as_date("2020-03-01") & start_date <= as_date("2021-02-28") ~ "1st Wave: Original",
    start_date >= as_date("2021-01-01") & start_date <= as_date("2021-06-30") ~ "2nd Wave: Alpha",
    start_date >= as_date("2021-07-01") & start_date <= as_date("2021-10-30") ~ "3rd Wave: Delta",
    start_date >= as_date("2021-12-01") & start_date <= as_date("2022-03-31") ~ "4th Wave: Omicron BA.1",
    start_date >= as_date("2022-04-01") & start_date <= as_date("2023-06-14") ~ "5th Wave: Omicron-BA",
    start_date >= as_date("2023-06-15") ~ "6th Wave:Omicron other",
    TRUE ~ NA_character_
  )) %>%
  filter(!is.na(wave))


#head(datdeaths_waves_v)




#Compute average death rates by region and wave
region_wave_death_rates_v <- datdeaths_waves_v %>%
  mutate(population = as.numeric(unlist(Population))) %>%
  group_by(region_name) %>%
  mutate(population = ifelse(is.nan(Population), NA, Population)) %>%
  fill(Population, .direction = "down")
region_data_v<- region_wave_death_rates_v %>%   
  group_by(region_name,wave) %>%
  summarise(
    total_deaths = sum(total_deaths, na.rm = TRUE),
    total_population = mean(Population, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(death_rate_per_100k = (total_deaths / total_population) * 100000)
#View(region_data)

library(ggplot2)



# Treand analysis of Death Rates Across COVID-19 Waves by Region

# Treat wave as an ordered factor (for display)
region_data_v <- region_data_v %>%
  mutate(wave = factor(wave, levels = unique(wave), ordered = TRUE))

plot_trend_v<- ggplot(region_data_v, aes(x = wave, y = death_rate_per_100k, group = region_name, color = region_name)) +
  geom_point(size = 3) +
  geom_line(aes(group = region_name)) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    title = "Linear Trend of Death Rates Across COVID-19 Waves by Region",
    x = "Wave",
    y = "Death Rate per 100,000",
    color = "Region"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

#plot_trend_v

#ggsave("docs/new/Linear Trend of Death Rates Across COVID-19 Waves by Region.pdf", plot= plot_trend, width = 12, height = 8)
#ggsave("docs/new/Linear Trend of Death Rates Across COVID-19 Waves by Region.png", plot = plot_trend, width = 12, height = 8, dpi = 300)



#death rates (per 100,000) by state and wave
state_wave_death_rates_v <- datdeaths_waves_v %>%
  mutate(population = as.numeric(unlist(Population))) %>%
  group_by(state_name) %>%
  mutate(population = ifelse(is.nan(Population), NA, Population)) %>%
  fill(Population, .direction = "down")
state_data_v <-state_wave_death_rates_v %>% 
  group_by(state_name,wave) %>% 
  summarise(
    total_deaths = sum(total_deaths, na.rm = TRUE),
    avg_population = mean(Population, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(death_rate_per_100k = (total_deaths / avg_population) * 100000)
#View(state_data)

# Table 2: top states by death rate per wave
topstate_by_wave_v <- state_data_v %>%
  group_by(wave) %>%
  slice_max(death_rate_per_100k, n = 1) %>%
  arrange(desc(death_rate_per_100k)) %>% 
  ungroup() 
summary_table1_v <- topstate_by_wave_v %>%
  select(State = state_name, Wave = wave,Population= avg_population, `Death Rate` = death_rate_per_100k) %>%
  arrange(Wave, desc(`Death Rate`))  
summary_table1_v <- topstate_by_wave_v %>%
  select(State = state_name, Wave = wave, `Death Rate` = death_rate_per_100k) %>%
  mutate(`Death Rate` = round(`Death Rate`, 1)) %>%
  arrange(Wave, desc(`Death Rate`))
kable(summary_table1_v, caption = "Table 1: Top States by COVID-19 Death Rate  per 100,000 population")

top5_by_wave_v <- state_data_v %>%
  group_by(wave) %>%
  slice_max(death_rate_per_100k, n = 5) %>%
  arrange(desc(death_rate_per_100k)) %>% 
  ungroup() 
# Prepare your summary table BEFORE using kable
summary_table_v <- top5_by_wave_v %>%
  select(State = state_name, Wave = wave, `Death Rate` = death_rate_per_100k) %>%
  arrange(Wave, desc(`Death Rate`))  # optional for nicer ordering
kable(summary_table_v, caption = "Top 5 States by COVID-19 Death Rate per Variant Wave")
summary_table_v <- top5_by_wave_v %>%
  select(State = state_name, Wave = wave, `Death Rate` = death_rate_per_100k) %>%
  mutate(`Death Rate` = round(`Death Rate`, 1)) %>%
  arrange(Wave, desc(`Death Rate`))

# Plot6: Top 5 States by COVID-19 Death Rate for Each Variant Wave facet by wave
plot6_v <- ggplot(top5_by_wave_v, aes(x = fct_reorder(state_name, death_rate_per_100k), y = death_rate_per_100k, fill = wave)) +
  geom_col(show.legend = FALSE) +
  coord_flip() +
  facet_wrap(~ wave, scales = "free_y") +
  labs(
    title = "Top 5 States by COVID-19 Death Rate for Each Variant Wave",
    x = "State",
    y = "Death Rate (per 100,000 people)"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    strip.text = element_text(face = "bold"),
    axis.text.x = element_text(angle = 0, hjust = 1)
  )
#plot6_v
#ggsave("docs/new/Top 5 States by COVID-19 Death Rate for Each Variant Wave.pdf", plot= plot6, width = 12, height = 8)
#2ggsave("docs/new/Top 5 States by COVID-19 Death Rate for Each Variant Wave.png", plot = plot6, width = 12, height = 8, dpi = 300)

```

```{r}
library(dplyr)
library(knitr)

top_state_per_year_v <- datdeaths_v %>%
  filter(mmwr_year != 2025) %>% 
  group_by(mmwr_year, state_name) %>%
  summarise(
    total_deaths = sum(total_deaths, na.rm = TRUE),
    total_population = mean(Population, na.rm = TRUE),
    death_rate_per_100k = (total_deaths / total_population) * 100000,
    .groups = "drop"
  ) %>%
  group_by(mmwr_year) %>%
  slice_max(death_rate_per_100k, n = 1) %>%  # top 1 state per year
  ungroup() %>%
  arrange(mmwr_year) %>%
  rename(
    Year = mmwr_year,
    State = state_name,
    Covid_Deaths = total_deaths,
    Population = total_population,
    Death_rate_per_100k = death_rate_per_100k
  ) %>%
  kable(caption = "Table 2: Summary - States with highest death rate (by year)")

top_state_per_year_v


top5_statesname_v <- datdeaths_v %>%
  filter(mmwr_year != 2025) %>% 
  group_by(mmwr_year, state_name) %>%
  summarise(
    total_deaths = sum(total_deaths, na.rm = TRUE),
    total_population = mean(Population, na.rm = TRUE),
    death_rate_per_100k = (total_deaths / total_population) * 100000,
    .groups = "drop"
  ) %>%
  group_by(mmwr_year) %>%
  slice_max(death_rate_per_100k, n = 1) %>%  # top 1 state per year
  ungroup() %>%
  slice_head(n = 5) %>%
  pull(state_name)
#top5_statesname_v

top5_data_v <- datdeaths_v %>%
  filter(state_name %in% top5_statesname_v) 


library(ggplot2)
top_state_per_year_data_v <- datdeaths_v %>%
  filter(mmwr_year != 2025) %>% 
  group_by(mmwr_year, state_name) %>%
  summarise(
    total_deaths = sum(total_deaths, na.rm = TRUE),
    total_population = mean(Population, na.rm = TRUE),
    death_rate_per_100k = (total_deaths / total_population) * 100000,
    .groups = "drop"
  ) %>%
  group_by(mmwr_year) %>%
  slice_max(death_rate_per_100k, n = 1) %>%  # top 1 state per year
  ungroup() %>% 
  arrange(mmwr_year) %>%
  rename(
    Year = mmwr_year,
    State = state_name,
    Covid_Deaths = total_deaths,
    Population = total_population,
    Death_rate_per_100k = death_rate_per_100k
  )
#top_state_per_year_data_v

plot8_v <- ggplot(top_state_per_year_data_v, aes(x = factor(Year), y = Death_rate_per_100k, fill = State)) +
  geom_col(width = 0.6) +
  labs(
    title = "Top State by COVID-19 Death Rate per 100,000 (by Year)",
    x = "Year",
    y = "Death Rate per 100,000",
    fill = "State"
  ) +
  theme_minimal(base_size = 20) +
  theme(legend.position = "bottom")
#plot8_v
#ggsave("docs/new/Top State by COVID-19 Death Rate per 100,000 (by Year).pdf", plot= plot8, width = 12, height = 8)
#ggsave("docs/new/Top State by COVID-19 Death Rate per 100,000 (by Year).png", plot = plot8, width = 12, height = 8, dpi = 300)
```

```{r}

# Plot9: Facet plot Covid deaths for top 10 states  
facet_plot_v<- ggplot(top5_data_v, aes(x = start_date, y = total_deaths)) +
  geom_line(color = "steelblue", size = 1) +
  geom_smooth(se = FALSE, color = "darkred", method = "loess", span = 0.3) +
  facet_wrap(~ state_name,scales = "fixed") +
  labs(
    title = "COVID-19 Deaths Over Time – Faceted by States",
    x = "Date",
    y = "Deaths"
  ) +
  scale_x_date(date_breaks = "9 months", date_labels = "%b\n%Y") +
  theme_minimal(base_size = 15) +
  theme(
    strip.text = element_text(face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
#facet_plot_v
#ggsave("docs/new/facet_top5_statesnew.png", plot = facet_plot, width = 12, height = 8, dpi = 300)
#ggsave("docs/new/facet_top5_statesnew.pdf", plot = facet_plot, width = 12, height = 8)


```

```{r}
library(ggplot2)
library(patchwork)

library(gridExtra)

myplot_v <- grid.arrange(plot2_v, plot4_v, plot_trend_v, plot6_v, ncol = 2)
myplot_v
#ggsave("docs/facet_top5_statesnew.png", plot = myplot, width = 12, height = 8, dpi = 300)
#ggsave("docs/facet_top5_statesnew.pdf", plot = myplot, width = 12, height = 8)
```

```{r}
library(dplyr)
library(ggplot2)
library(lubridate)
library(tidyverse)
library(broom)
library(knitr)
library(MMWRweek)
library(gganimate)
library(httr2)
library(jsonlite)
get_cdc_data_IK <- function(endpoint) {
  api_url <- paste0("https://data.cdc.gov/resource/", endpoint, ".json")
  data <- request(api_url) |> 
    req_url_query("$limit" = 10000000) |> 
    req_perform() |> 
    resp_body_json(simplifyVector = TRUE)
  return(data)
}

deaths_raw_IK <- get_cdc_data_IK("r8kw-7aab")
vax_raw_IK <- get_cdc_data_IK("rh2h-3yt2")
```

## Pre-Processing and Data Exploration

```{r}
covid_deaths_IK <- deaths_raw_IK %>%
  mutate(week_ending_date = as.Date(week_ending_date),
         start_date = as.Date(start_date),
         end_date = as.Date(end_date)) %>%
  filter(group == "By Week", !is.na(total_deaths), !state %in% c("United States", "Puerto Rico", "District of Columbia", "New York City"),!is.na(covid_19_deaths)) %>% 
  select(-c(group, month, data_as_of, footnote))
```

## Expected Excess Deaths Calculation

```{r}
excess_death <- covid_deaths_IK %>% 
  mutate(mmwr_week = as.integer(mmwr_week), 
         total_deaths = as.numeric(total_deaths), 
         covid_19_deaths = as.numeric(covid_19_deaths), 
         percent_of_expected_deaths = as.numeric(percent_of_expected_deaths)) %>% 
  mutate(expected_deaths = total_deaths/(percent_of_expected_deaths/100), 
         excess_deaths = total_deaths - expected_deaths, 
         percent_change = (excess_deaths/expected_deaths)*100) %>% 
  filter(!is.na(expected_deaths))

#excess_death
```

## Data Visualizations

### Excess Death Over Time/State & Against COVID-19 Waves

```{r}
excess_death <- excess_death %>% 
  mutate(region = state.region[match(state, state.name)])

regions_loess <- ggplot(excess_death, aes(x = week_ending_date, y = percent_change, color = region)) + 
  geom_line(alpha = 0.4) + 
  labs(title = "Excess Deaths (%) by Week and Region", y = "% Excess Mortality", x = "Week") +  
  geom_smooth(se = FALSE, method = "loess", span = 0.3, color = "black") + 
  theme_minimal()

regions_loess 

loess_line <- ggplot(excess_death, aes(x = week_ending_date, y = percent_change)) +
  geom_smooth(se = FALSE, method = "loess", span = 0.3, color = "black")

variant_waves_IK <- tibble::tibble(
  start = as_date(c("2020-03-01", "2021-01-01", "2021-07-01", "2021-12-01", "2022-04-01", "2023-06-15")),
  end   = as_date(c("2021-02-28", "2021-06-30", "2021-10-30", "2022-03-31", "2023-06-14", "2024-03-31")),
  variant = c("1st Wave: Original", 
              "2nd Wave: Alpha", 
              "3rd Wave: Delta", 
              "4th Wave: Omicron BA.1", 
              "5th Wave: Omicron Other",
              "6th Wave: XBB")
)

# Adding rectangles for each wave
ymin_val <- min(excess_death$percent_change, na.rm = TRUE)
ymax_val <- max(excess_death$percent_change, na.rm = TRUE)
for (i in 1:nrow(variant_waves_IK)) {
  loess_line <- loess_line +
    annotate("rect",
             xmin = variant_waves_IK$start[i],
             xmax = variant_waves_IK$end[i],
             ymin = ymin_val,
             ymax = ymax_val,
             alpha = 0.2,
             fill = scales::hue_pal()(nrow(variant_waves_IK))[i])
}

loess_overlay <- loess_line +
  labs(title = "Excess Mortality Trend by Week with Variant Waves", x = "Week", y = "% Excess Mortality") +
  theme_minimal()

loess_overlay
```

### Smoothing Line with Animation

```{r}
loess_fit_IK <- loess(percent_change ~ as.numeric(week_ending_date), data = excess_death, span = 0.3)
excess_death$loess_smooth <- predict(loess_fit_IK)

# gg_animate <- ggplot(excess_death, aes(x = week_ending_date)) + geom_rect(data = variant_waves_IK, aes(xmin = start, xmax = end, ymin = -Inf, ymax = Inf, fill = variant), inherit.aes = FALSE, alpha = 0.2) + geom_line(aes(y = loess_smooth), color = "black", linewidth = 1) +
# labs(title = "Excess Mortality Trend with Variant Waves", x = "Week", y = "% Excess Mortality", fill = "Variant Waves") + theme(axis.text.x = element_text(angle = 45, hjust = 1),plot.title = element_text(hjust = 0.5)) + transition_reveal(week_ending_date) + theme_minimal()
# 
# animate(gg_animate, width = 900, height = 600, fps = 20, duration = 10, renderer = gifski_renderer("loess_animation.gif"), loop = FALSE)

```

### States Most Frequently / Least Frequently in Top 5 for Excess Deaths

```{r}
mean_excess_yr <- excess_death %>%
  mutate(mmwr_year = epiyear(week_ending_date)) %>%
  group_by(mmwr_year, state) %>%
  summarize(mean_percent_change = 100 * mean(excess_deaths) / mean(expected_deaths), .groups = "drop")

top_5_yr <- mean_excess_yr %>%
  group_by(mmwr_year) %>%
  slice_max(mean_percent_change, n = 5, with_ties = FALSE) %>%
  ungroup() %>%
  group_by(state) %>%
  summarize(years_in_top5 = n(), top_years = paste(sort(unique(mmwr_year)), collapse = ", "))

top_5_yr <- top_5_yr %>%
  mutate(top_years_wrapped = gsub("^((?:[^,]+, ){2}[^,]+),\\s*", "\\1\n", top_years))

top_bar5 <- top_5_yr %>%
  slice_max(years_in_top5, n = 5, with_ties = FALSE) %>%
  ggplot(aes(x = reorder(state, years_in_top5), y = years_in_top5)) +
  geom_col(fill = "steelblue") +
  geom_text(aes(label = top_years_wrapped), hjust = -0.1, size = 3) +
  labs(title = "States Most Frequently in the Top 5 for % Excess Mortality (2020-2025)",
       x = "State", y = "Years in Top 5") +
  coord_flip() +
  theme_minimal() +
  ylim(0, max(top_5_yr$years_in_top5) + 1)
top_bar5

lower_5_year <- mean_excess_yr %>%
  group_by(mmwr_year) %>%
  slice_min(mean_percent_change, n = 5, with_ties = FALSE) %>%
  ungroup() %>%
  group_by(state) %>%
  summarize(years_in_low5 = n(), bottom_years = paste(sort(unique(mmwr_year)), collapse = ", "))

lower_5_year <- lower_5_year %>%
  mutate(bottom_years_wrapped = gsub("^((?:[^,]+, ){2}[^,]+),\\s*", "\\1\n", bottom_years))

lower_bar5 <- lower_5_year %>%
  slice_max(years_in_low5, n = 5, with_ties = FALSE) %>%
  ggplot(aes(x = reorder(state, years_in_low5), y = years_in_low5)) +
  geom_col(fill = "steelblue") +
  geom_text(aes(label = bottom_years_wrapped), hjust = -0.1, size = 3) +
  labs(title = "States Most Frequently in the Bottom 5 for % Excess Mortality (2020-2025)",
       x = "State", y = "Years in Bottom 5") +
  coord_flip() +
  theme_minimal() +
  ylim(0, max(lower_5_year$years_in_low5) + 1)
lower_bar5
```

### Relationship Between Observed and Expected Deaths, and COVID-19 Deaths

```{r}
ggplot(excess_death, aes(x = covid_19_deaths, y = excess_deaths)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "blue") +
  labs(title = "Relationship Between Reported COVID-19 Deaths and Excess Mortality",
       x = "Reported COVID-19 Deaths", y = "Excess Mortality (# of Deaths)") +
  theme_minimal()

covid_excess_death <- summary(lm(excess_deaths ~ covid_19_deaths, data = excess_death))
covid_excess_death


```

### How Many Excess Deaths Can Be Attributed to COVID-19?

```{r}
diff_death_IK <- excess_death %>%
  mutate(diff_death = excess_deaths - covid_19_deaths) %>%
  filter(excess_deaths >= 0)

linreg_diff <- diff_death_IK %>%
  ggplot(aes(x = covid_19_deaths, y = diff_death)) +
  geom_point() +
  geom_hline(yintercept = 0, color = "red") +
  labs(x = "Deaths by COVID-19", y = "Difference in Excess and COVID-19 Deaths") +
  theme_minimal()
linreg_diff


```

### Impact of Respiratory Diseases on Excess Deaths

```{r}
excess_death <- excess_death %>%
  mutate(year = case_when(
    year %in% c("2019/2020", "2019-2020") ~ 2020,
    year %in% c("2020/2021", "2020-2021") ~ 2021,
    year %in% c("2021/2022", "2021-2022") ~ 2022,
    TRUE ~ as.numeric(year)
  )) %>% filter(!is.na(year))

vax_clean_IK <- vax_raw_IK %>%
  select(date, series_complete_daily, location, mmwr_week) %>%
  mutate(series_complete_daily = as.numeric(series_complete_daily),
         state = state.name[match(location, state.abb)],
         year = epiyear(date)) %>%
  group_by(year, mmwr_week, state) %>%
  summarize(date = max(as.Date(date)),
            series_complete_weekly = sum(series_complete_daily, na.rm = TRUE),
            .groups = "drop") %>%
  mutate(mmwr_week = as.integer(mmwr_week))

vax_excess <- left_join(excess_death, vax_clean_IK, by = c("year", "mmwr_week", "state")) %>%
  mutate(
    pneumonia_deaths_alone = as.numeric(pneumonia_deaths) - as.numeric(pneumonia_and_covid_19_deaths),
    influenza_deaths = as.numeric(influenza_deaths)
  ) %>%
  filter(year < 2024)

other_excess_death <- summary(lm(
  excess_deaths ~ covid_19_deaths + pneumonia_deaths_alone + influenza_deaths +
    series_complete_weekly +
    I(pneumonia_deaths_alone * covid_19_deaths) + I(influenza_deaths * covid_19_deaths),
  data = vax_excess
))
#other_excess_death

other_excess_death_tidy <- tidy(other_excess_death)

relabelled_table <- other_excess_death_tidy %>%
  select(term, estimate, p.value) %>%
  mutate(
    term = recode(term,
      `(Intercept)` = "Intercept",
      `covid_19_deaths` = "Deaths by Covid-19",
      `pneumonia_deaths_alone` = "Deaths by Pneumonia Only",
      `influenza_deaths` = "Influenza Deaths",
      `series_complete_weekly` = "Vaccination Series Complete",
      `pneumonia_deaths_alone*covid_19_deaths` = "Deaths by Pneumonia Only * Deaths by Covid-19",
      `I(influenza_deaths*covid_19_deaths)` = "Influenza Deaths * Deaths by Covid-19"
    )
  ) %>%
  kable(
    caption = "Regression Results: Predicting Excess Deaths",
    col.names = c("Term", "Estimate", "P-Value"),
    format = "html"
  )
relabelled_table


```

### Pre-Covid Respiratory Death Trends

```{r}
resp_precovid <- read.csv("data/NCHSData15.csv") %>%
  rename(
    year = Year,
    mmwr_week = Week,
    pneumonia_deaths = Pneumonia.Deaths,
    influenza_deaths = Influenza.Deaths
  ) %>%
  filter(year < 2020)

future_weeks <- resp_precovid %>%
  group_by(mmwr_week) %>%
  summarize(expected_resp_deaths = mean(pneumonia_deaths, na.rm = TRUE)) %>%
  ungroup() %>%
  crossing(year = 2020:2025)

ggplot(resp_precovid, aes(x = mmwr_week, y = pneumonia_deaths, color = factor(year))) +
  geom_point() +
  theme_minimal() +
  labs(x = "Week", y = "Pneumonia Deaths", color = "Year")

ggplot() +
  geom_point(data = resp_precovid, aes(x = mmwr_week, y = pneumonia_deaths, color = as.factor(year)), alpha = 0.5, shape = 1) +
  geom_point(data = future_weeks %>% mutate(year = "2020-2025"), aes(x = mmwr_week, y = expected_resp_deaths, color = factor(year))) +
  theme_minimal() +
  labs(title = "Expected Pneumonia Deaths (2020–2025) with Historical Baseline",
       x = "Week", y = "Expected Pneumonia Deaths", color = "Year")

```

### Excess Respiratory Deaths + COVID = Excess Death?

```{r}
resp_analysis <- excess_death %>%
  mutate(
    pneumonia_alone = as.numeric(pneumonia_deaths) - as.numeric(pneumonia_and_covid_19_deaths),
    resp_deaths = pneumonia_alone
  ) %>%
  group_by(mmwr_week, year) %>%
  summarize(
    pneumonia_deaths = sum(resp_deaths, na.rm = TRUE),
    covid_19_deaths = sum(covid_19_deaths, na.rm = TRUE),
    excess_deaths_total = sum(excess_deaths, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  left_join(future_weeks, by = c("year", "mmwr_week")) %>%
  mutate(
    diff_resp = pneumonia_deaths - expected_resp_deaths,
    excess_resp_only = ifelse(diff_resp > 0, diff_resp, 0),
    covid_plus_resp = covid_19_deaths + excess_resp_only,
    residual_covid_only = excess_deaths_total - covid_19_deaths,
    residual_covid_plus_resp = excess_deaths_total - covid_plus_resp,
    residual_difference = 100 * (residual_covid_only - residual_covid_plus_resp) / residual_covid_only
  )

#excess deaths not attributed to covid 
excess_notcovid <- ggplot(resp_analysis, aes(x = covid_19_deaths, y = residual_covid_plus_resp)) +
  geom_point() +
  labs(x = "Covid-19 Deaths", y = "Excess Pneumonia Mortality") +
  theme_minimal() +
  geom_hline(yintercept = 0, color = "red")
excess_notcovid

#excess deaths not attributed to covid or pneumonia 
excess_pneum <- ggplot(resp_analysis, aes(x = MMWRweek2Date(MMWRyear = year, MMWRweek = mmwr_week))) +
  geom_line(aes(y = residual_covid_only, color = "COVID only"), linewidth = 1) +
  geom_line(aes(y = residual_covid_plus_resp, color = "COVID + Pneumonia"), linewidth = 1.3, linetype = "dashed") +
  geom_hline(yintercept = 0, linetype = "dotted") +
  labs(y = "Unexplained Excess Deaths", x = "Week", color = "Model",
       title = "Explained vs Unexplained Excess Deaths: Adding Pneumonia") +
  theme_minimal()
excess_pneum

### Expected deaths vs Excess Deaths over Time 
exp_v_unexp <- ggplot(resp_analysis, aes(x = MMWRweek2Date(MMWRyear = year, MMWRweek = mmwr_week))) +
  geom_line(aes(y = residual_difference), color = "red", size = 1.2) +
  geom_hline(yintercept = 0, linetype = "dotted", color = "black") +
  labs(title = "Impact of Including Pneumonia Deaths on Excess Mortality Estimates",
       x = "Week", y = " % Unexplained Excess Deaths Accounted For") +
  theme_minimal()
exp_v_unexp

```
